Subglacial and subaerial fluvial sediment transport capacity respond differently to water discharge variations

### 2024-07-01 Delaneyetal2024.tex
ESurf update

%% Copernicus Publications Manuscript Preparation Template for LaTeX Submissions
%% ---------------------------------
%% This template should be used for copernicus.cls
%% The class file and some style files are bundled in the Copernicus Latex Package, which can be downloaded from the different journal webpages.
%% For further assistance please contact Copernicus Publications at: production@copernicus.org
%% https://publications.copernicus.org/for_authors/manuscript_preparation.html

@@

\newcommand{\aref}[1]{\textbf{Reference #1}}
\newcommand{\afig}[1]{\textbf{Figure #1}}
\newcommand{\TODO}[1]{\textbf{TODO: \color{red}#1}}
\newcommand{\ian}[1]{{\textbf{\color{blue}Ian sayz:} \color{blue} #1} }
\newcommand{\mauro}[1]{{\textbf{\color{green}Mauro says:} \color{green} #1} }
\newcommand{\alpine}{\textit{ALPINE}}
\newcommand{\icesheet}{\textit{ICESHEET}}
\newcommand{\m}{$\,\mathrm{m}$\,}
\newcommand{\cm}{$\,\mathrm{cm}$\,}
\newcommand{\mma}{$\,\mathrm{mm  \, a^{-1}}$\,}
\newcommand{\mmma}{$\,\mathrm{m^3\, a^{-1}}$\,}
\newcommand{\mmms}{$\,\mathrm{m^3\, s^{-1}}$\,}
%\newcommand{\unit}[1]{$\mathrm{#1}$}

@@

% \Author[affil]{given_name}{surname}

@@

\affil[1]{Institut des dynamiques de la surface terrestre (IDYST), Universit\'{e} de Lausanne, B\^{a}timent G\'{e}opolis, 1015 Lausanne, Switzerland}
\affil[2]{Department of Geosciences, University of Fribourg, Ch. du Musée 1700, Fribourg, Switzerland}
\affil[3]{Laboratory of Hydraulics, Hydrology and Glaciology (VAW), ETH-Z\"urich, H\"onggerbergring 26, 8093 Z\"urich, Switzerland}
\affil[4]{Swiss Federal Institute for Forest, Snow and Landscape Research (WSL) Z\"uricherstrasse 111, 8903 Birmensdorf, Switzerland}
%% The [] brackets identify the author with the corresponding affiliation. 1, 2, 3, etc. should be inserted.

%% If an author is deceased, please mark the respective author name(s) with a dagger, e.g. "\Author[2,$\dag$]{Anton}{Smith}", and add a further "\affil[$\dag$]{deceased, 1 July 2019}".

%% If authors contributed equally, please mark the respective author names with an asterisk, e.g. "\Author[2,*]{Anton}{Smith}" and "\Author[3,*]{Bradley}{Miller}" and add a further affiliation: "\affil[*]{These authors contributed equally to this work.}".

@@

%% These dates will be inserted by Copernicus Publications during the typesetting process.

@@

\begin{abstract}
  Sediment transport capacity in both subaerial and subglacial channels depends on the shear stress exerted across the channel bottom, which varies with water velocity and channel width.
  In subaerial channels, water discharge variations are accommodated by flow depth and width changes, along with water velocity.
  However, in subglacial channels,  water discharge changes mainly result in velocity variations, as the channel geometry evolves slowly (over days).
  Here, we present relationships between water discharge and sediment transport under several channel types.
  They demonstrate that sediment transport is highly non-linear with respect to water discharge with fixed subglacial channel size, and in fact, can very less than subaerial flow if water discharge changes slowly enough to be in equilibrium with subglacial channel size.
  We then apply hydrographs from an Alpine glacier and the Greenland Ice Sheet to subglacial and subaerial hydraulics models, revealing greater variability in the inferred sediment transport capacity in subglacial than in subaerial channels, as subglacial channels evolve in size.
  This variability stems from the slowly evolving channel size that forces water velocity to accommodate variations in water discharge.
  Furthermore, the changing channel size means that sediment transport capacity peaks before the maximum water discharge.
  Results also show that high subglacial sediment transport rates and water velocities can occur across a wide range of water discharges, meaning that in a transport-limited system, there could be a variable relationship between sediment and water discharge.
  These findings are discussed in the context of understanding sediment discharge from glaciers with different hydro-climatic forcings and assumptions of channel shape.
  We also examine the implications for interpreting records of sediment discharge in glacierized catchments
\end{abstract}

@@

Changes in glacier dynamics and hydrology have prompted numerous recent studies of sediment transport processes in cold regions \citep[e.g.][]{li2022,vergara2022,zhang2022}.
Increases in sediment transport have been observed in Greenland \citep{bendixen2017}, the European Alps \citep{costa2017}, the Himalayas \citep{li2021}, and the Andes \citep{vergara2022}.
To explain observed changes to sediment transport in glacierized catchments requires examining the processes controlling sediment discharge and its variations with water discharge  \citep[e.g.][]{riihimaki2005,swift2005}.

Periods of millennia or longer, processes such as glacier abrasion and quarrying sculpt landscapes and create sediment to be transported fluvially from under glaciers \citep[c.f.][]{hallet1979,iverson2012,ugelvig2018}.
Pressurized subglacial water can transport this sediment from underneath glaciers \citep{walder1994,creyts2013,beaud2018,delaney2019} if it is reachable by the water.

In a transport-limited regime, sediment discharge is controlled by sediment transport capacity, defined as the amount of sediment the water can carry.
In both subglacial and subaerial channels, sediment transport capacity depends on the shear stress between water and the sediment it flows over \citep{shields1936,meyer1948,engelund1967} along with the width of the channel bottom $w$ over which sediment mobilizes.
The shear stress $\tau$ responds to the velocity of water $v$ flowing through the channel so that
\begin{linenomath*}
  \begin{equation}
    \label{eq:tau}
    \tau \propto v^2.
  \end{equation}
\end{linenomath*}
% 
Following mass conservation, the mean velocity of the water flowing through a  channel is
\begin{linenomath*}
  \begin{equation}
    \label{eq:v}
    v = \frac{Q}{S},
  \end{equation}
\end{linenomath*}
where $Q$ is water discharge,  and $S$ is the channel's wetted area.

In subaerial channels, operating with open channel flow, the wetted area $S$ of the channel evolves with changing water discharge $Q$, by changing both the channel's width and the water depth \citep{leopold1953}.
The change in water depth results in a proportional increase in water velocity, and the shear stress $\tau$ increases according to Equation~\ref{eq:tau}.

The response of water velocity to changing water discharge in subglacial channels differs from subaerial ones.
The size of subglacial channels is controlled by the opposing processes of channel opening by frictional heating of water flow on the one hand versus creep closure by ice flow on the other \citep{rothlisberger1972}.
As a result, the subglacial channel size only evolves relatively slowly over days, whereas water discharge can vary more quickly over hours \citep[e.g.][]{iken1986,andrews2014,nanni2020}.
Therefore, on short periods, subglacial water flow behaves more like pipe flow, and changes in water discharge $Q$ are mainly accommodated by changing water velocity $v$ \citep[Equation~\ref{eq:v} and Figure~\ref{fig:cartoon}; ][]{alley1997}.

@@

Sediment mobilization in subaerial and subglacial channels, therefore, responds differently to changing water discharge.
These differences are implicitly included in a range of models quantifying sediment transport in both subglacial and subaerial channels  \citep[e.g.][]{walder1994,alley1997,tucker1997,creyts2013,beaud2018,delaney2019,hewitt2019,wickert2019}.
To date, explicit parameterizations of subglacial sediment transport capacity with respect to water discharge have assumed fixed channel size and demonstrate a very non-linear response in sediment transport capacity to water discharge \citep{alley1997}.
The modeling frameworks examining subglacial sediment transport with evolving channel size minimally discuss the impact of different hydrological forcing on sediment transport capacity to understand the result of hydro-climatic conditions on subglacial sediment dynamics \citep{creyts2013,beaud2018,delaney2019,hewitt2019}.
Yet, sediment discharge capacity controls the mobilization and deposition of sediment, making it one of the most fundamental aspects of subglacial sediment evacuation, especially over short timescales.

Despite the physical differences between the subaerial and subglacial systems, water discharge and sediment export are often compared in contemporary glacial environments \citep[e.g.][]{willis1996,pearce2003,richards2003,swift2005,chu2009,tedstone2012,chu2012,overeem2017,delaney2018,swift2021,lu2022,andresen2024}.
These comparisons are natural given that the majority of sediment evacuated from glaciers over contemporary seasonal timescales is done so fluvially \citep[e.g.][]{collins1989, riihimaki2005,delaney2024}.
In subaerial systems, water discharge and sediment transport capacity for a grain size are closely linked \citep[e.g. ][]{schmidt2006,pitlick2021} \ian{check citiation}.
However, the potential for a highly variable response in sediment transport capacity in subglacial sediment transport may affect the interpretation of sediment transport records in glacial systems \citep[e.g.][]{ganti2016,mancini2023}, if properties associated with subaerial systems are assumed.
This relationship should be clarified as sediment dynamics change in glacierized regions along with changing hydrology, increased variability of water discharge \citep{lane2019} and extreme glacier melt  \citep[e.g.][]{cremona2023,overeem2015} that may drive high magnitude sediment transport events.

This manuscript has two objectives: 1) evaluate the variability of subglacial sediment transport capacity with evolving subglacial channel size and different hydrographs and 2) establish if sub-seasonal variations in water discharge can co-vary with sediment transport capacity in subglacial systems.
First, to reinforce the highly variable behavior of subglacial sediment transport capacity,  we present algebraic formulations of the sediment transport capacity response to water discharge in subaerial, pipe-flow, and equilibrium R-channel conditions, extending the relationships presented in \citet{alley1997}.
We then use numerical models to examine sediment transport capacity in subglacial channels that evolve in size and subaerial channels.
They are then applied to proglacial hydrological records from an Alpine glacier in Switzerland (Fieschergletscher) and a land-terminating glacier in Greenland (Leverett Glacier).
We run models with an ensemble of parameters to evaluate the sediment transport capacity variability in subglacial channels compared to subaerial ones.
Lastly, model outputs demonstrate the particular processes that can drive variability in sediment discharge capacity in subglacial systems.
Together, these findings indicate differences in the relationships between water discharge, channel geometry, and water velocity in the subglacial and subaerial channels.
The manuscript then uses the identified differences between the subaerial and subglacial channel types to discuss the implications for interpreting the processes driving variability in sediment transport records and the impact of these differences on sediment export from glaciers.

@@

Water discharge data is used from Alpine and ice sheet settings collected downstream of the glacier, assuming no water storage in the proglacial area.
The Alpine site (labeled \alpine{}) is  Fieschergletscher in the Swiss Alps ($46^\circ\,29'\,07''$ N, $8^\circ\,08'\,3''$ E).
The water discharge data used here was collected at a $1$\unit{min} interval from May 24, 2014, to October 10, 2014 \citep[Figure~\ref{fig:Qw}\,a][]{felix2022}.

@@

Note that we omit analysis of suspended sediment discharge records due to the role of sediment supply in capturing variations in sediment discharge, in addition to sediment transport capacity \citep[e.g.][]{delaney2019}.

\begin{figure}[hbt!]
  \centering
  \includegraphics[width=0.8\linewidth]{Qw_plot.pdf}
  \caption{Water discharge from Fieschergletcher (\alpine) and Leverett Glacier (\icesheet).
    Note differences in variability and quantity of water discharge over the season.
  }
  \label{fig:Qw}
\end{figure}

\section{Methods}
\label{sect:meth}
The two models described below (Sections ~\ref{sect:sub_mode}~and~\ref{sect:fluv}) represent relationships amongst water discharge, water velocity, and channel geometry in both subaerial and subglacial channels (Table~\ref{table:vpm}).
Both models use the measured discharge to calculate water velocity, shear stress, and width-integrated shear stress, upon which both suspended sediment and bedload transport depend \citep[Figure \ref{fig:cartoon}; ][]{shields1936}.
Our evaluation in terms of shear stress omits the selection of a sediment transport relationship and a grain-size parameter \citep[e.g.][]{shields1936,meyer1948}.

@@

To evaluate the shear stress of water flowing across sediments underneath a glacier, the subglacial channel model accounts for the channel geometry and the water's velocity.
To do this, we use a lumped hydraulics model from \citet{werder2010b}, itself based upon  \citet{clarke1996}.

Here, it is assumed that the water is transported through a subglacial channel \citep[Figure~\ref{fig:cartoon}; ][]{rothlisberger1972} beneath a glacier with channel length $l$, with a flatbed and a mean thickness of $h_{ice}$.
The channel size grows from melt due to frictional heating from water flow and closes due to the creep of the ice.
The formulation here does not consider the englacial storage of water.
The evolution of subglacial channel size $S_g$ is given as
\begin{linenomath*}
  \begin{equation}
    \label{eq:dS_dt}
    \frac{\partial S_g}{\partial t} = C_1 \frac{Q \Delta h}{l} - C_2 \left(h_{o}-\frac{\Delta h}{2}\right)^n\,S_g,
  \end{equation}
\end{linenomath*}
\noindent where $t$ is time, $C_1= (1-\rho_wc_pc_t)\,\frac{\rho_wg}{\rho_iL}$ and $C_2=2A(\frac{\rho_wg}{n})^n$ are constants (complete values given in Table~\ref{table:vpm}), $g$ is the acceleration due to gravity, $Q$ is water discharge, $\Delta h$ is the hydraulic potential change over $l$, $h_{o}= \frac{\rho_i}{\rho_w} h_{ice}$ is the mean ice overburden pressure expressed in meter water equivalent ($\rho_w$ is density of water; $\rho_i$ is density of ice), and $n$ is Glen's n \citep{glen1955} (usually $n=3$).
The first term on the equation's right side represents the opening of the channel through frictional heating, while the following term represents the creep closure of the channel from ice deformation.

@@

The hydraulics parameterization presented in \citet{tucker1997} is implemented to represent the subaerial channel.
Using mass conservation and the Darcy-Weisbach relationship as well as assuming that the channel is sufficiently wide compared to its depth so that the hydraulic radius is well approximated by the flow depth,
the shear stress $\tau_f$ at the river bed is
\begin{linenomath*}
  \begin{equation}
    \label{eq:DW_tau}
    \tau_f=\frac{\rho_w\,g^{\frac{2}{3}}\,f_f^{\frac{1}{3}}}{2}\, \Big(\frac{Q}{w_f} \Big)^{\frac{2}{3}} \,\nabla z_c^{\frac{2}{3}},
  \end{equation}
\end{linenomath*}
where $\nabla z_c$ is the channel slope, and $f_f$ is the friction factor for subaerial channels \citep{tucker1997}.
Channel width $w_f$ is
\begin{equation}
  \label{eq:wcf}
  w_f = k \, Q^{\alpha},
\end{equation}
% 
where $k$ is a constant and $\alpha=\frac{1}{3}$ is a commonly chosen exponent \citep{leopold1953}.
Also following the Darcy-Weisbach, subaerial water velocity, $v_f$, is given as
\begin{equation}
  \label{eq:vf}
  v_f = \sqrt{\frac{8\,\tau_f}{f_f\,\rho_w}}.
\end{equation}
% 
As above, the width-integrated shear stress is $w_f\tau_f$.

@@

\begin{table}[hbt!]
  \centering
  \caption{Variables, parameters, and constants used in this work.
    Where two values are given, the first refers to  \alpine{}, a scenario from Fieschergletscher, and the second to a glacier marginal to the \icesheet{} a scenario from Leverett Glacier. If a second line is given, it refers to the range of values examined in the parameter search}
  \small 
  \begin{tabular}{ l  c  c c }
    Name &Symbol&  Value&Units \\
         && (\alpine{} or \icesheet{})\\
    \hline
    \textbf{Variables}  & & & \\
    Water discharge  & $Q$& & $\mathrm{m^{3}\,s^{-1}}$ \\
    Water velocity (subglacial, subaerial)  & $v$, ($v_g,\,v_{f}$)& & $\mathrm{m\,s^{-1}}$ \\
    Channel wetted area (subglacial, subaerial) &  $S_g, S_f$& & $\mathrm{m^2}$     \\
    Channel depth (subaerial) & $H$&& $\mathrm{m}$\\
    Hydraulic diameter &$D_h$&&$\mathrm{m}$\\
    Width of channel floor (subglacial, subaerial) & $w$, ($w_g,w_f$)&  & $\mathrm{m}$     \\
    Hydraulic head &$\Delta h$&& $\mathrm{m}$\\
    Hydraulic gradient &$\Psi=\frac{\Delta h}{l}$&& $\mathrm{m\, m^{-1}}$\\
    Shear stress (subglacial, subaerial) & $\tau$, ($\tau_g,\,\tau_f$) && $\mathrm{Pa \, m^{-2}}$ \\
    Stream power & $\Omega$ && $\mathrm{ kg \, m\, s^{-3}}$ \\

@@

The models above are applied to proglacial discharge records from the Fieschergletscher (scenario \alpine{}) and the Leverett Glacier (scenario \icesheet{}).
Outputs of the models represent generalizable sediment transport characteristics from these hydrographs, rather than actual hydraulic conditions.
To generalize these scenarios, \alpine{}  is exemplified by Fieschergletscher's relatively thin ice thickness \citep[$h_{ice}$= $225$\,\unit{m}][]{grab2021}, low water discharge ($\sim\,10$\,\unit{m}$^3$\,\unit{s}$^{-1}$) and high diurnal variability in water discharge (Figure~\ref{fig:Qw}\, b).
\icesheet{}  is exemplified by Leverett glacier's  thick ice  \citep[$h_{ice}$= $700$\,\unit{m}; ][]{morlighem2017}, high water discharge ($\sim\,300$\,\unit{m}$^3$\,\unit{s}$^{-1}$)  and low diurnal variability in water discharge (Figure~\ref{fig:Qw}\, b).

The first experiment aims to characterize the variability in sediment discharge capacity in subglacial compared to subaerial channels across a range of channel slopes and shapes, along with friction factors.
Additionally, we examine the effects of different water discharge variability by smoothing the water discharge over different periods.
The results below present water velocity ($v_g$, $v_f$), shear stress ($\tau_g$, $\tau_f$), and width-integrated shear stress ($w_g\tau_g$, $w_f\tau_f$) from the subglacial and subaerial models.
For brevity, these values together are referred to as ``model outputs'' in the text.
The subglacial model outputs are generally larger than their subaerial equivalents due to their large hydraulic gradients, thus greater sediment transport capacity would be expected in subglacial channels compared to subaerial ones \citep{alley1997}.
To evaluate variability, we subtract the model outputs from their values smoothed using daily averages,  detrending the seasonal signal (Figure~\ref{fig:Qw}).
We then calculate the variability in the model output by taking the standard deviation of this detrended time series.

>>>>

%%Copernicus Publications Manuscript Preparation Template for LaTeX Submissions
%% ---------------------------------
%% This template should be used for copernicus.cls
%% The class file and some style files are bundled in the Copernicus Latex Package, which can be downloaded from the different journal webpages.
%% For further assistance please contact Copernicus Publications at: production@copernicus.org
%% https://publications.copernicus.org/for_authors/manuscript_preparation.html

@@

\newcommand{\aref}[1]{\textbf{Reference #1}}
\newcommand{\afig}[1]{\textbf{Figure #1}}
\newcommand{\TODO}[1]{\textbf{TODO: \color{red}#1}}
\newcommand{\ian}[1]{{\textbf{\color{blue}Ian sayz:} \color{blue} #1} }
\newcommand{\mauro}[1]{{\textbf{\color{green}Mauro says:} \color{green} #1} }
\newcommand{\alpine}{\textit{ALPINE}\,}
\newcommand{\icesheet}{\textit{ICESHEET}\,}
\newcommand{\m}{$\,\mathrm{m}$\,}
\newcommand{\cm}{$\,\mathrm{cm}$\,}
\newcommand{\mma}{$\,\mathrm{mm  \, a^{-1}}$\,}
\newcommand{\mmma}{$\,\mathrm{m^3\, a^{-1}}$\,}
\newcommand{\mmms}{$\,\mathrm{m^3\, s^{-1}}$\,}
% \newcommand{\unit}[1]{$\mathrm{#1}$}

@@

@@

\affil[1]{Institut des dynamiques de la surface terrestre (IDYST), Universit\'{e} de Lausanne, B\^{a}timent G\'{e}opolis, 1015 Lausanne, Switzerland}
\affil[2]{Department of Geosciences, University of Fribourg, Ch. du Musée 1700, Fribourg, Switzerland}
\affil[3]{Laboratory of Hydraulics, Hydrology and Glaciology (VAW), ETH-Z\"urich, H\"onggerbergring 26, 8093 Z\"urich, Switzerland}
\affil[4]{Swiss Federal Institute for Forest, Snow and Landscape Research (WSL) Z\"uricherstrasse 111, 8903 Birmensdorf, Switzerland}

@@

@@

\begin{abstract}
  Sediment transport capacity in both subaerial and subglacial channels depends on the shear stress exerted across the channel bottom.
  This quantity varies with water velocity and channel width.
  In subaerial channels,  sediment transport capacity can evolve consistently because water discharge variations are accommodated by changes in flow depth, width, and water velocity.
  By contrast, subglacial channels grow at a slower rate than most water discharge variations, causing a variable relationship between channel shape, water velocity, and discharge.
  Using numerical models applied to two observed hydrographs and formulations, we evaluate the differences in sediment transport capacity variability between subaerial and subglacial channels.
  We also establish the conditions under which subglacial sediment transport capacity could co-vary with water discharge.
  The formulations of hydraulics and numerical models confirm that subglacial sediment transport is highly non-linear with respect to water discharge.
  Yet, subglacial sediment transport capacity  can vary less if water discharge variations are closer to equilibrium with subglacial channel size.
  In another numerical experiment, we show that the changing channel size results in sediment transport capacity peaking before the maximum water discharge.
  Results also indicate that high subglacial sediment transport capacities can occur across a wide range of water discharges.
  This implies that  hysteresis in channel size causes a variable relationship between sediment and water discharge in a transport-limited subglacial system.
  The implications are discussed for sediment discharge from glaciers with different hydro-climatic forcings and assumptions of channel shape.
  These findings can also improve interpretations of sediment discharge records in glacierized catchments.
\end{abstract}

@@

Changes in glacier dynamics and hydrology have motivated numerous recent studies of sediment transport processes in cold regions \citep[e.g.][]{li2022,vergara2022,zhang2022}.
Increases in sediment transport have been observed in Greenland \citep{bendixen2017}, the European Alps \citep{costa2017}, the Himalayas \citep{li2021}, and the Andes \citep{vergara2022}.
To accurately explain observed changes to sediment transport in glacierized catchments requires examining the processes controlling sediment discharge and its variations with water discharge  \citep[e.g.][]{riihimaki2005,swift2005}.

Glacier abrasion and quarrying sculpt landscapes and create sediment to be transported fluvially from under glaciers over periods of millennia or longer \citep[c.f.][]{hallet1979,iverson2012,ugelvig2018}.
Pressurized subglacial water can transport this sediment from underneath glaciers \citep{walder1994,creyts2013,beaud2018,delaney2019} if it is reachable by the water.

In a transport-limited regime, sediment discharge is controlled by sediment transport capacity, defined as the amount of sediment the water can carry.
In both subglacial and subaerial channels, sediment transport capacity depends on the shear stress between water and the sediment it flows over \citep{shields1936,meyer1948,engelund1967}, along with the width of the channel bottom $w$ over which sediment mobilizes.
The shear stress $\tau$ responds to the velocity of water $v$ flowing through the channel so that
\begin{linenomath*}
  \begin{equation}
    \label{eq:tau}
    \tau \propto v^2.
  \end{equation}
\end{linenomath*}
% 
Following mass conservation, the mean velocity of the water flowing through a  channel is
\begin{linenomath*}
  \begin{equation}
    \label{eq:v}
    v = \frac{Q}{S},
  \end{equation}
\end{linenomath*}
where $Q$ is water discharge,  and $S$ is the channel's wetted area.

In subaerial channels operating with open channel flow, the wetted area $S$ of the channel evolves with changing water discharge $Q$, by changing both the channel width and the water depth \citep{leopold1953}.
The change in water depth results in a proportional increase in water velocity, and the shear stress $\tau$ increases according to Equation~\ref{eq:tau}.

The response of water velocity to changing water discharge in subglacial channels differs from subaerial ones, however.
The size of subglacial channels is controlled by the opposing processes of channel opening by frictional heating of water flow on the one hand versus creep closure by ice flow on the other \citep{rothlisberger1972}.
As a result  the subglacial channel size only evolves relatively slowly over days, whereas water discharge can vary more quickly over hours \citep[e.g.][]{iken1986,andrews2014,nanni2020}.
Therefore, subglacial water flow behaves more like pipe flow over short periods (hours, days).
Changes in water discharge $Q$ are mainly accommodated by changing water velocity $v$ \citep[Equation~\ref{eq:v} and Figure~\ref{fig:cartoon}; ][]{alley1997}.

@@

Sediment mobilization in subaerial and subglacial channels, therefore, responds differently to changing water discharge.
These differences are implicitly included in a range of models quantifying sediment transport in both subglacial and subaerial channels  \citep[e.g.][]{walder1994,alley1997,tucker1997,creyts2013,beaud2018,delaney2019,hewitt2019,wickert2019}.
To date, explicit parameterizations of subglacial sediment transport capacity with respect to water discharge have assumed fixed channel size  \citep{alley1997}. and
This result demonstrates a strongly non-linear response in subglacial sediment transport capacity to water discharge.
Several modeling frameworks examining subglacial sediment transport with evolving channel size \citep{creyts2013,beaud2018,delaney2019,hewitt2019}.
However, these works minimally discuss the impact of different hydrologies on sediment transport capacity underneath glaciers for interpreting sediment transport records.
Understanding these processes is imperative for establishing the effect of hydro-climatic conditions on subglacial sediment dynamics.
Particularly because sediment discharge capacity controls the mobilization and deposition of sediment, making it a fundamental aspect of subglacial sediment evacuation, especially over short timescales.

Despite the physical differences between the subaerial and subglacial systems, water discharge and sediment export are often compared in contemporary glacierized catchments \citep[e.g.][]{willis1996,pearce2003,richards2003,swift2005,chu2009,tedstone2012,chu2012,overeem2017,delaney2018,swift2021,lu2022,andresen2024}.
Many of theses studies discuss the disparity between water discharge and sediment export, highlighting the non-linear relationship between the two, along with the role of sediment access or production.
Yet, a variable relationship between water discharge and sediment export could be expected from glaciers with evolving conduit size in a transport-limited, or hydraulically controlled, regime.
The response of sediment transport capacity in subglacial sediment transport may affect the interpretation of sediment transport records in these glacierized catchments systems \citep[e.g.][]{ganti2016,mancini2023}.
It is important that this relationship is clarified as sediment dynamics change in glacierized regions along with changing hydrology \citep[e.g.][]{brunner2019}, increasing variability of water discharge \citep{lane2019} and occurrence of extreme glacier melt  \citep[e.g.][]{overeem2015,cremona2023} that may drive high magnitude sediment transport events.

This manuscript has two objectives: 1) to evaluate the variability of subglacial sediment transport capacity with evolving subglacial channel size and different hydrographs and 2) to establish whether sub-seasonal water discharge can co-vary with sediment transport capacity in subglacial systems.
First, to demonstrate subglacial sediment transport capacity's highly variable nature,  we present algebraic formulations of the sediment transport capacity response to water discharge in subaerial, pipe-flow, and equilibrium R-channel conditions.
These formulations extend the relationships presented in \citet{alley1997}.
We then use numerical models to examine sediment transport capacity in subglacial channels that evolve in size and subaerial channels.
They are applied to proglacial hydrological records from an Alpine glacier in Switzerland (Fieschergletscher) and a land-terminating glacier in Greenland (Leverett Glacier).
We run a model ensemble to evaluate the sediment transport capacity variability in subglacial channels compared to subaerial channels.
Model outputs demonstrate the specific processes that can drive variability in sediment discharge capacity in subglacial systems.
These findings indicate differences in the relationships amongst water discharge, channel geometry, and water velocity in subglacial and subaerial channels.
The manuscript then discusses the implications of variability in sediment transport capacity from glaciers and the interpretation of sediment transport records.

@@

Water discharge data is  from Alpine and ice sheet settings collected downstream of the glacier, and assumes no water storage in the proglacial area.
The Alpine site (labeled \alpine{}) is  Fieschergletscher in the Swiss Alps ($46^\circ\,29'\,07''$ N, $8^\circ\,08'\,3''$ E).
The water discharge data used here was collected at a $1$\unit{min} interval from May 24, 2014, to October 10, 2014 \citep[Figure~\ref{fig:Qw}\,a][]{felix2022}.

@@

Note that we omit analysis of suspended sediment discharge records due to their dependence on sediment supply in capturing variations in sediment discharge, in addition to sediment transport capacity \citep[e.g.][]{delaney2019}.

\begin{figure}[hbt!]
  \centering
  \includegraphics[width=0.8\linewidth]{Fig2.pdf}
  \caption{Water discharge from Fieschergletcher (\alpine) and Leverett Glacier (\icesheet).
    Note differences in variability and quantity of water discharge over the season.
  }
  \label{fig:Qw}
\end{figure}

\section{Methods}
\label{sect:meth}
The two models described below (Sections ~\ref{sect:sub_mode}~and~\ref{sect:fluv}) represent relationships amongst water discharge, water velocity, and channel geometry in both subaerial and subglacial channels (Table~\ref{table:vpm}).
Both models use the measured discharge to calculate water velocity, shear stress, and width-integrated shear stress, upon which both suspended sediment and bedload transport depend \citep[Figure \ref{fig:cartoon}; ][]{shields1936}.
Our choice to evaluate shear stress omits selecting a sediment transport relationship and a grain-size parameter \citep[e.g.][]{shields1936,meyer1948}.

@@

To evaluate the shear stress of water flowing across sediments underneath a glacier, the subglacial channel model accounts for the channel geometry and the water's velocity.
We use a lumped hydraulics model from \citet{werder2010b}, itself based upon  \citet{clarke1996}.

Here, it is assumed that the water is transported through a subglacial channel \citep[Figure~\ref{fig:cartoon}; ][]{rothlisberger1972} beneath a glacier with channel length $l$, with a flatbed and a mean thickness of $h_{ice}$.
The channel size grows from melt due to frictional heating from water flow and closes due to ice creep.
The formulation here does not consider the englacial storage of water.
The evolution of subglacial channel size $S_g$ is given as
\begin{linenomath*}
  \begin{equation}
    \label{eq:dS_dt}
    \frac{\partial S_g}{\partial t} = C_1 \frac{Q \Delta h}{l} - C_2 \left(h_{o}-\frac{\Delta h}{2}\right)^n\,S_g,
  \end{equation}
\end{linenomath*}
\noindent where $t$ is time, $C_1= (1-\rho_wc_pc_t)\,\frac{\rho_wg}{\rho_iL}$ and $C_2=2A(\frac{\rho_wg}{n})^n$ are constants (values in Table~\ref{table:vpm}), $g$ is the acceleration due to gravity, $Q$ is water discharge, $\Delta h$ is the hydraulic potential change over $l$, $h_{o}= \frac{\rho_i}{\rho_w} h_{ice}$ is the mean ice overburden pressure expressed in meter water equivalent ($\rho_w$ is density of water; $\rho_i$ is density of ice), and $n$ is Glen's n \citep{glen1955} (usually $n=3$).
The first term on the equation's right side represents the channel opening by frictional heating, while the following term represents channel closure from ice deformation.

@@

The hydraulics parameterization presented in \citet{tucker1997} is implemented to represent the subaerial channel.
This model uses mass conservation and the Darcy-Weisbach relationship and assumes that the channel is sufficiently wide compared to its depth.
This latter assumption means that the hydraulic radius is well approximated by the flow depth.
Therefore, the resulting shear stress $\tau_f$ at the river bed is
\begin{linenomath*}
  \begin{equation}
    \label{eq:DW_tau}
    \tau_f=\frac{\rho_w\,g^{\frac{2}{3}}\,f_f^{\frac{1}{3}}}{2}\, \Big(\frac{Q}{w_f} \Big)^{\frac{2}{3}} \,\nabla z_c^{\frac{2}{3}},
  \end{equation}
\end{linenomath*}
where $\nabla z_c$ is the channel slope, and $f_f$ is the friction factor for subaerial channels \citep{tucker1997}.
Channel width $w_f$ is
\begin{equation}
  \label{eq:wcf}
  w_f = k \, Q^{\alpha},
\end{equation}
% 
where $k$ is a constant and $\alpha=\frac{1}{3}$ is a commonly chosen exponent \citep{leopold1953}.
Also following the Darcy-Weisbach, subaerial water velocity, $v_f$, is given as
\begin{equation}
  \label{eq:vf}
  v_f = \sqrt{\frac{8\,\tau_f}{f_f\,\rho_w}}.
\end{equation}
% 
As above, the width-integrated shear stress is $w_f\tau_f$.

@@

\begin{table}[hbt!]
  \centering
  \caption{Variables, parameters, and constants used in this work.
    Where two values are given, the first refers to  \alpine{}, a scenario from Fieschergletscher, and the second to a glacier marginal to the \icesheet{} a scenario from Leverett Glacier.
    A second line is given refers to the range of values examined in the parameter search.}
  \small 
  \begin{tabular}{ l  c  c c }
    Name &Symbol&  Value&Units \\
         && (\alpine{} or \icesheet{})\\
    \hline
    \textbf{Variables}  & & & \\
    Water discharge  & $Q$& & $\mathrm{m^{3}\,s^{-1}}$ \\
    Water velocity (subglacial, subaerial)  & $v$, ($v_g,\,v_{f}$)& & $\mathrm{m\,s^{-1}}$ \\
    Channel wetted area (subglacial, subaerial) &  $S_g, S_f$& & $\mathrm{m^2}$     \\
    Channel depth (subaerial) & $H$&& $\mathrm{m}$\\
    Hydraulic diameter &$D_h$&&$\mathrm{m}$\\
    Width of channel floor (subglacial, subaerial) & $w$, ($w_g,w_f$)&  & $\mathrm{m}$     \\
    Hydraulic head &$\Delta h$&& $\mathrm{m}$\\
    Hydraulic gradient &$\Psi=\frac{\Delta h}{l}$&& $\mathrm{m\, m^{-1}}$\\
    Shear stress (subglacial, subaerial) & $\tau$, ($\tau_g,\,\tau_f$) && $\mathrm{Pa \, m^{-2}}$ \\
    Stream power & $\Omega$ && $\mathrm{ kg \, m\, s^{-3}}$ \\

@@

The models above are applied to proglacial discharge records from the Fieschergletscher (scenario \alpine{}) and the Leverett Glacier (scenario \icesheet{}).
The model outputs represent generalizable sediment transport characteristics from these hydrographs, rather than actual hydraulic conditions.
To generalize these scenarios, \alpine{}  is exemplified by relatively thin ice thickness \citep[$h_{ice}$= $225$\,\unit{m}][]{grab2021}, low water discharge ($\sim\,10$\,\unit{m}$^3$\,\unit{s}$^{-1}$) and high diurnal variability in water discharge at Fieschergletscher (Figure~\ref{fig:Qw}\, b).
\icesheet{}  is exemplified by   thick ice  \citep[$h_{ice}$= $700$\,\unit{m}; ][]{morlighem2017}, high water discharge ($\sim\,300$\,\unit{m}$^3$\,\unit{s}$^{-1}$)  and low diurnal variability in water discharge at Leverett glacier (Figure~\ref{fig:Qw}\, b).

The first experiment aims to characterize the variability in sediment discharge capacity in subglacial relative to subaerial channels across a range of channel slopes and shapes, along with friction factors.
Additionally, we examine differing water discharge variability by smoothing the water discharge over different period from $15$\,\unit{min} up to $15$\,\unit{days}.
The results below present water velocity ($v_g$, $v_f$), shear stress ($\tau_g$, $\tau_f$), and width-integrated shear stress ($w_g\tau_g$, $w_f\tau_f$) from the subglacial and subaerial models.
For brevity, these values together are referred to as ``model outputs'' in the text.
To evaluate variability, we subtract the model outputs from their values smoothed using daily averages,  detrending the seasonal signal.
We then calculate the variability in the model output by taking the standard deviation of this detrended time series.
